---
toc: true
toc-expand: 2
editor: 
  markdown: 
    wrap: 72
---

::: callout-note
## Credits & copyright

BDA ASCA author: "Johan Westerhuis"
:::

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Install and load the necessary packages

ASCA will be applied using the HDANOVA package, which also needs the
mixlm package.

```{r}
# One-time installation

if (!requireNamespace("mixlm", quietly = TRUE)) {
	devtools::install_github("khliland/mixlm")
}
if (!requireNamespace("HDANOVA", quietly = TRUE)) {
	devtools::install_github("khliland/HDANOVA")
}

# in newer version of R you can also use pakage manager pak to install these packages
# pak::pak("khliland/mixlm")
# pak::pak("khliland/HDANOVA")

# load the libraries
library(mixlm)
library(HDANOVA)
```

# Exercise during lecture

```{r}
# create a simple dataset with 3 groups and 12 samples

# create response variable y and two different design matrices for the same grouping, one with reference coding and one with sum coding
y = c(5,6,7,6,3,2,4,3,1,2,3,1)
X0 = c(1,1,1,1,1,1,1,1,1,1,1,1) # not needed in calculation; Intercept is automatically calculated in lm function

# create design matrices for reference coding and sum coding
Xr1 = matrix(c(0,0,0,0,1,1,1,1,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1),ncol = 2) # 
Xs1 = matrix(c(-1,-1,-1,-1,1,1,1,1,0,0,0,0,-1,-1,-1,-1,0,0,0,0,1,1,1,1),ncol = 2)

# make linear models with reference coding and sum coding
bref = lm(y~Xr1)
bsum = lm(y~Xs1)

# look at the summary of both models
summary(bref)
summary(bsum)
```

::: {.callout-tip icon="false"}
## Question: What are the differences in interpretation?

What are the estimates for $\mu$, $\alpha_1$, $\alpha_2$, $\alpha_3$ in
both cases?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   For Reference coding $\mu = 6$, $\alpha_1 = 0$, $\alpha_2 = -3$,
    $\alpha_3 = -4.25$.
-   Reference coding forces the intercept to be equal to the groupmean
    of group 1. The a1 value is 0 as group 1 is the reference group.
-   For Sum coding $\mu = 3.5833$, $\alpha_1 = 2.4166$,
    $\alpha_2 = -0.5833$, $\alpha_3 = -1.8333$. The $\alpha_1 = 2.4166$
    value can be obtained as the sum of $\alpha_1$, $\alpha_2$ and
    $\alpha_3$ has to be 0. Sumcoding makes the intercept to be equal to
    the average of all groups. The interpretation of the effects is
    therefore different between Reference and Sum coding. Note however,
    that the full model statistics are exactly equal.
:::

# Brassicaceae study

In this exercise you will use the **HDANOVA** and **mixlm** packages to
explore linear models and ASCA for analysis of high dimensional data
from an designed study setup.

Plants in the Cabbage family (Brassicaceae) make very specific defense
compounds, called glucosinolates, when under attack by herbivores. There
are about 120 different versions of these glucosinolates, but here we
measure only 11. To simulate a herbivore attack and elicit the plant to
make glucosinolates, the plant hormone jasmonic acid (JA) was applied to
either the roots (Root induced) or to the leaves (Shoot induced) of
cabbage plants (B. oleracea). The glucosinolate levels for 11 different
glucosinolates were measured at 1, 3, 7 and 14 days after treatment.
This measurement was destructive so that for each time-point different
plants were analyzed. For each treatment, 5 different plants were
measured at each time point. Besides the two treatments there was also a
control group that received no JA treatment. We would like to know how
the different treatments affect the glucosinolate composition and
whether the glucosinolate composition varies with time.

```{r}
# load data and design
Data = read.csv("Data.csv")
Design = read.csv("F.csv")
```

In the Environment window (top right) click on Data and Design to see
the content of these two sets. Data contains the measurements, and
Design the experimental settings of the samples.

::: {.callout-tip icon="false"}
## Question: Is the design balanced?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

Yes, each combination of Treatment and Time has 5 replicates.
:::

::: {.callout-tip icon="false"}
## Question: What are important specifics you can observe by a quick look at the Data?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

Some variables have many zeros, some variables have much higher values
than others.
:::

## Data Preprocessing

First make a boxplot of the data to see which preprocessing steps should
be taken.

```{r}
boxplot(Data)
```

::: {.callout-tip icon="false"}
## Question: Wat problematic issues do you observe for which variables?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

| Some features have few extreme large values (e.g. Neo, GBC), their distribution seems skewed to the right. Others have zero's (feature 2, 6, 9 etc).

Sizes of the different features are also different (GBC and NEO are
large).
:::

In such a case a square root transform is often a good choice to
preprocess the data.

::: {.callout-tip icon="false"}
## Question: Explain why a square root transform is a useful transformation for these data.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

| If there is more variation between the replicates for higher values of a feature (the feature distribution is rights skewed), then a transformation can correct that.

If there are zero values, a log transform is problematic. A square root
transform does not have problems with zero values, and makes the
distributions more symmetric around the mean. A square root transform
also makes the sizes of the features more equal.
:::

Make a data frame containing factors Time and Treatment and Data.

```{r}
# Make a data frame containing factors Time and Treatment and Data
dat = data.frame(sqrt(Data),Design);
dat$Time = as.factor(dat$Time)
dat$Treatment = as.factor(dat$Treatment)
```

## Linear models and ANOVA

To explore the effect of the treatments on the glucosinolates, let's
start with an ANOVA model for a single response using the lm function.
By default the lm function uses SUM coding for the treatment levels.

```{r}
# make 1-way ANOVA model for PRO using Treatment as a factor
mod_Treatment = lm(PRO~Treatment, data = dat)
print(anova(mod_Treatment))
```

There is clearly a difference between the treatments as the p-value is
very low. The summary command wil provide some more info on the
treatment levels, and provide residual plots to check the assumptions on
the residuals.

```{r}
# Get a little bit more information
summary(mod_Treatment)
plot(mod_Treatment, which = 1:2)
```

::: {.callout-tip icon="false"}
## Question: Explain why the effect of one of the treatment levels is not calculated. How can you obtain information about this level?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

If all levels are represented in the design matrix, and the intercept is
estimated as well, then there are too many free parameters leading to a
situation with infinite number of solutions. By forcing the extra
restriction that all treatment effect add up to 0 (which is the result
of Sum coding) this situation is solved. The shoot level effect, which
is not estimated, thus equals 0.05008.
:::

::: {.callout-tip icon="false"}
## Question: What is your interpretation of this model? What can you say about the root, shoot and control levels? How can you interpret the residual plots?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   The intercept is the average value of all measurements.
-   The root treatment provides PRO levels on average 0.24 lower than
    the average, while the shoot treatment provides PRO levels 0.19
    higher than the average.
-   The control level is not indicated, but the sum of all treatment
    levels equals zero, thus the control level provides PRO levels
    around 0.05 higher than the average.
-   The root and shoot levels are clearly different from zero according
    to the t-test.
-   The residual vs Fitted plot shows residuals are nicely spread around
    zero for all fitted values (which are 1.12 - 0.24 for control;
    1.12+0.05 for control and 1.12+0.19 for shoot).
-   The QQ residuals plot to check normality shows only three samples
    with slightly higher residuals than expected. This shows that the
    residuals follow a normal distribution quite well.
:::

### Two-way crossed ANOVA

Instead of a single factor we also want to know the effect of both
factors and also their interaction.

```{r}
# two-way crossed ANOVA
mod <- lm(GBN ~ Time*Treatment, data = dat) # Note the * means all main effects and interactions
print(anova(mod))
```

For GBN, we see that both factors (Time and Treatment) are significant
as is their interaction. Also here check the significance of the levels
of each of the factors and interaction using the summary command.

```{r}
summary(mod)
plot(mod, which = 1:2)
```

::: {.callout-tip icon="false"}
## Question: Explain your interpretation of this model.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   For GBN there is a clear treatment effect and also some timepoints
    are different.
-   There is a single interaction clearly different from zero.
-   The GBN values at time 3 for the control treatment are 0.17 higher
    than could be expected from their main factor results control
    treatment effect and day 3 effect.
:::

Different codings can be used for the factors. Here you can compare sum
coding and treatment (reference) coding.

```{r}
# Different codings
mod_sum <- lm(GBN ~ Time*Treatment, data = dat, contrasts = "contr.sum")
print(anova(mod_sum))
summary(mod_sum)

mod_treat <- lm(GBN ~ Time*Treatment, data = dat, contrasts = "contr.treatment")
print(anova(mod_treat))
summary(mod_treat)
```

::: {.callout-tip icon="false"}
## Question: Discuss the differences between the two types of coding. When is one preferred over the other?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

The difference is mainly what is considered the intercept (base level)
of the model. If a control level is present, one would like to see how
the treated level is different from the control instead of how the both
differ from their average.
:::

::: {.callout-tip icon="false"}
## Question: What is the interpretation of the intercept in the models? What is the meaning of the estimated coefficients in each model?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

In the SUM coded model, the intercept represents the overall mean, but
in the treatment coding the intercept represents the mean of the Control
treatment samples at time 1. Check this! The effects of the other levels
mean the difference between those levels and the Control at time 1
level.
:::

## ASCA

ASCA generalizes ANOVA or the linear model to the multivariate setting.
The models are applied to all response variables, and the variation due
to the different factors (and their interactions) is summarized into
effect matrices. Then a PCA model can be applied to each effect matrix
separately of in combination to find variation that is systematic over
all the variables. This variation is then summarized with PCA scores and
loadings.

We will use the **HDANOVA** package to apply asca. The HDANOVA package
expects to be in the long format. The next sections transforms the data
from wide to long format. Furthermore, the factors Time and Treatment
have to be defined as factors.

```{r}
# Reshape data from wide to long format
Metabolites = data.frame(Time = dat$Time,Treatment = dat$Treatment, Metab = I(as.matrix(dat[,1:11])))
Metabolites$Time = as.factor(Metabolites$Time)
Metabolites$Treatment = as.factor(Metabolites$Treatment)
```

Then make an ASCA model of the metabolite data using both Time and
Treatment as factors together with their interaction.

```{r}
# ASCA model of all variables simultaneously
mod_asca <- asca(Metab ~ Time*Treatment, data=Metabolites)
summary(mod_asca)
```

Open de mod_asca list and look at all the parameters that have been
calculated. Click the scores section. The scores are calculated For each
of the effect matrices. This is also the case for the loadings and
Projected scores. The latter are the score values calculated with back
projection, using the residuals and the effect matrices together, as
discussed in the lecture.

```{r}
# Simple functions to use
scoreplot(mod_asca, factor="Treatment", legend = ("topright"), main="Treatment scores", projections = FALSE)
scoreplot(mod_asca, factor="Treatment", legend = ("topright"), main="Projected Treatment scores")
scoreplot(mod_asca, factor="Treatment", legend = ("topright"), main="Treatment scores Spider plot", spider = TRUE)
loadingplot(mod_asca, factor="Treatment", scatter=TRUE, labels="names", main="Treatment loadings")


# For more control you can obtain information from the mod_asca list, e.g. when you only want to plot the top 8 loadings
# Loading plot for only important loadings

Var_Exp_Treatment = round(100 * mod_asca[["singulars"]][["Treatment"]]^2 / sum(mod_asca[["singulars"]][["Treatment"]]^2))
P_Treatment = mod_asca[["loadings"]][["Treatment"]];
Importance = P_Treatment[,1]^2 + P_Treatment[,2]^2;
NIMP = 8; # number of important variables
sorted_Importance = sort(Importance, decreasing = TRUE, index.return = TRUE)
Importance_Index = sorted_Importance$ix[1:NIMP]
plot(P_Treatment[Importance_Index,1],P_Treatment[Importance_Index,2],type = "n",xlab=paste0("Comp 1 (", Var_Exp_Treatment[1], "%)"), ylab=paste0("Comp 2 (", Var_Exp_Treatment[2], "%)"), main = "Important Treatment loadings")
text(P_Treatment[Importance_Index,1],P_Treatment[Importance_Index,2],colnames(Data[,Importance_Index]))

```

::: {.callout-tip icon="false"}
## Question: What is your interpretation of the ASCA model of the Treatment factor?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   There is a clear difference between the shoot and control
    treatments, which are described by the 1^st^ PC of the treatment
    factor. GBC and NEO have the largest contribution to this effect.
-   The 2^nd^ PC which is much smaller wrt explained variation shows how
    the root treatment is different from control and shoot, and here GBN
    and PRO have the largest contribution. Check this by looking in the
    data whether this observation is correct!
:::

It is important to check whether the factors and interactions are
significant. This can be done using permutation testing.

```{r}
mod.perm <- asca(Metab ~ Time*Treatment, data=Metabolites, permute = TRUE)
summary(mod.perm)
par(mfrow=c(1,3))
permutationplot(mod.perm, factor = "Time")
permutationplot(mod.perm, factor = "Treatment")
permutationplot(mod.perm, factor = "Time:Treatment")
```

::: {.callout-tip icon="false"}
## Question: What conclusion can you draw from the permutation test results?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

Both factors as well as the interaction have a significant effect on the
glucosinolate levels. For the main effects none of the permutations gave
better results than the true grouping, while for the interaction only
few permutations were larger than the true grouping.
:::

#### Combining Main and Interaction effects

Sometimes it makes sense to combine effects, e.g., a main factor and an
interaction. This is especially the case when the effect of Treatments
are studied over Time. Here we combine the effects of Treatment and the
Time:Treatment interaction. Note the effect matrices of Treatment and
Time:Treatment are estimated independently, but the PCA is applied on
the sum of these two effect matrices.

```{r}
mod.comb <- asca(Metab ~ Time + comb(Treatment + Time:Treatment), data=Metabolites)
summary(mod.comb)
```

If combined effects have a Time factor, the scores can be plotted
against the time using the timeplot function. For this plot the legend
command has to be called separately.

```{r}
par(mfrow=c(1,2), mar = c(4,4,1,1))
timeplot(mod.comb, factor="Treatment", time="Time", comb=2, comp=1,  x_time=TRUE)
legend("topright",col = 1:3,lty=1:3,levels(Metabolites$Treatment))
loadingplot(mod.comb, factor = "Treatment+Time:Treatment", comp=1)
```

::: {.callout-tip icon="false"}
## Question: Give your interpretation of this combined effect.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

In the scoreplot we see the combined effect of Treatment and
Time:Treatment interaction. This shows that the effect of the treatments
is not constant over time. The biological explanation is that both root
and shoot treatments have an effect, but after day 3, the root effect
disappears and the glucosinolate levels go back to the control values.
The shoot effect however is consistent and even increases over time.
Variables 8 (GBC) and 11 (NEO) are largely effected by these treatment
differences.
:::

Finally, it can be useful to also explore the residual matrix to see
whether there are effects in the data that is not modeled for.

```{r}
# Make score and loading plots and interpret.
par(mfrow=c(1,2)) 
loadingplot(mod.comb, factor="Residuals", scatter=TRUE, labels="names", main="Residual loadings")
scoreplot(mod.comb, factor="Residuals", main="Residual scores")
```

::: {.callout-tip icon="false"}
## Question: Why is it useful to also explore the residuals and interpret the PCA model of the residuals.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

We can see that some plants (the cluster bottom right deviate from the
main group. These will have higher values of NEO and lower values of GBC
than the other plants.)
:::

# Additional exercises

In this 2^nd^ exercise you have to do some coding yourself. Look at the
code provided above to see how to use the functions.

### The MoTrPAC data

The Molecular Transducers of Physical Activity Consortium (MoTrPAC)
profiled the dynamic change of multiple omics layers in several organs
of rats undergoing physical training. The results are presented in the
Nature article "Temporal dynamics of the multi-omic response to
endurance exercise training".
<https://www.nature.com/articles/s41586-023-06877-w> The data was
obtained from their website MoTrPAC-data.org.

### Experimental design

Six-month-old male and female Fischer 344 rats were subjected to
progressive treadmill endurance exercise training for 1, 2, 4 or 8
weeks. Sex-matched sedentary, untrained rats were used as controls (0).
In this application we will focus on metabolomics data from plasma.
Unfortunately these measurements do not all come from the same rats.
Note that Females have Sex = 1 and Males have Sex = 2.

#### Load and view the data

Load the metabolomics_Values (Data) and the metabolomics_Exp (Design)
files. For visualization reasons we will only select the lipids subset.

As preprocessing we decided to not use a transformation of the data, but
only use meancentering and scaling by the standard deviation of the
lipid.

```{r}
# load data and design

df_metabolomics_Values = read.csv("Plasma_Metabolomics_Values_ab.csv", row.names=1)
df_metabolomics_Exp = read.csv("Plasma_Metabolomics_Exp_ab.csv", row.names=1)

df_metabolomics_Lipids = df_metabolomics_Values[,80:613]

# make boxplots of the data before preprocessing
par(mfrow=c(1,3)) 
boxplot(df_metabolomics_Lipids)

# Preprocess the data by centering and scaling 
CData = scale(df_metabolomics_Lipids, center = TRUE, scale = FALSE)
SData = scale(df_metabolomics_Lipids, center = TRUE, scale = TRUE)

# Make boxplots of the data after centering and scaling
boxplot(CData)
boxplot(SData)
```

::: {.callout-tip icon="false"}
## Question: Why did we preprocess the lipids data this way?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   From the 1^st^ boxplot we see that the Lipids data is not centered.
-   We can also see the distributions are not skewed but rather
    symmetrical around their mean, thus transformation is not necessary.
-   Scaling seems necessary as the ranges of the centered data are
    rather different for some metabolites.
:::

#### Data handling

Just as above the data has to be made in the long format to be able to
use the **HDANOVA** package. Make sure the factors Sex and Group are
defined as factors.

```{r }
# make the data frame in long format
Data_Values_long = data.frame(Sex = as.factor(df_metabolomics_Exp$Rat.Sex), Training = as.factor(df_metabolomics_Exp$Group), Metabolites = I(as.matrix(SData)))
```

#### ASCA model of main factors:

Create an ASCA model using the only the main factors Sex and Training.
Do not consider interaction terms yet.

```{r }
# ASCA model of main factors
mod_asca1 <- asca(Metabolites ~ Sex + Training, data=Data_Values_long)

# Look at the summary of the model
summary(mod_asca1)
```

::: {.callout-tip icon="false"}
## Question: Explain the meaning of Sum.Sq and Expl.var. What is your interpretation of the ASCA model?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   Sum.Sq is the sum of squares of the effect matrix, calculated by
    squaring each element in the effect matrix and then adding all these
    values.
-   Expl.var is the percentage explained variation from the total
    variation in the Metabolites data. It is calculated by dividing the
    Sum.Sq of a factor (e.g. Sex) by the Sum.Sq of the original data
    (Metabolites).
-   There is a strong effect for factor Sex and a smaller effect for
    factor Training.
:::

#### Visualization

Make score plots of the Training effect

::: {.callout-tip icon="false"}
## Question: Show a score plot of the Training effect.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

```{r}
# Write your code for a score plots of the Training effect.
scoreplot(mod_asca1, factor="Training", legend = ("bottomleft"), main="Projected Treatment scores")

```
:::

Below the loading plots of PC1 vs PC2 of the Training factor is given.
For visualization reasons we only show the most important variables.

```{r}
# Loading plot of only 20 most important metabolites
# determine explained variance per principal component 
Var_Exp_Training = 100*(mod_asca1[["singulars"]][["Training"]]^2/sum(mod_asca1[["singulars"]][["Training"]]^2))
Var_Exp_Training = round(Var_Exp_Training,0)
# extract loadings of the Training factor
P_Training = mod_asca1[["loadings"]][["Training"]];
# combine the loadings of PC1 and PC2 into an importance value
Importance = sqrt(P_Training[,1]^2 + P_Training[,2]^2);
NIMP = 7; # number of important variables
# sort the importance values and get the index of the top 20 most important variables
sorted_Importance = sort(Importance, decreasing = TRUE, index.return = TRUE)
Importance_Index = sorted_Importance$ix[1:NIMP]
# plot the most important variables in the loading plot
plot(P_Training[Importance_Index,1],P_Training[Importance_Index,2],xlab=paste0("Comp 1 (", Var_Exp_Training[1], "%)"), ylab=paste0("Comp 2 (", Var_Exp_Training[2], "%)"))
text(P_Training[Importance_Index,1],P_Training[Importance_Index,2],colnames(df_metabolomics_Lipids[,Importance_Index]))
colnames(df_metabolomics_Lipids[,Importance_Index])
```

::: {.callout-tip icon="false"}
## Question: Give your interpretation of the score and loading plots. \[1\]
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

The scoreplot shows the group average of the 5 training groups by the
solid marker, while the individual rats are indicated with the open
markers. Component 1 mainly makes a distinction between Training week1
and week2 on the right and the control (0), 4 and 8 weeks training
samples are mainly on the left part of the plot. PC2 makes a distinction
between Control and 8 weeks of training. The training groups are not
well separated, however the upper left part seems to be mainly related
to 8 weeks of training.

The upper left extreme loadings (LPC17.1.0.0_feature2 and
LPC14.0.0.0_andLPC0.0.14.0_feature1and2) both have high values for 8
weeks of training.
:::

#### ASCA model with interaction

::: {.callout-tip icon="false"}
## Question: Make a new model in which the two-factor interaction Sex:Training is also included in the model.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

```{r}
# Write the code for an ASCA model including main factors and interaction
mod_asca_Int <- asca(Metabolites ~ Sex*Training, data=Data_Values_long)
# Provide the summary of the model
summary(mod_asca_Int)
```
:::

::: {.callout-tip icon="false"}
## Question: What is your conclusion of this ASCA model with interaction?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

| The interaction explains an additional 7.7% of the variation in the data.

This means that the Time effect is somewhat different for males compared
to females.
:::

#### Statistical Inference

To conclude whether the main factors and interaction are significant,
apply a permutation test and plot the permutation result using the
permutation plot function for both main factors and the interaction.

```{r}
# Permutation test for the ASCA model with interaction
mod_asca_Int_perm <- asca(Metabolites ~ Sex + Training + Sex:Training, data=Data_Values_long, permute = TRUE)
summary(mod_asca_Int_perm)
# make permutation plots for the main factors and the interaction
par(mfrow=c(1,3))
permutationplot(mod_asca_Int_perm, factor = "Sex")
permutationplot(mod_asca_Int_perm, factor = "Training")
permutationplot(mod_asca_Int_perm, factor = "Sex:Training")

```

::: {.callout-tip icon="false"}
## Question: Explain why in ASCA a permutation test is used and not an F-test.
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

The F-statistic assumes normally distributed data and also requires the
degrees of freedom for each effect. We have not tested the normality of
the data, and degrees of freedom is unclear in the multivariate
situation.
:::

::: {.callout-tip icon="false"}
## Question: What is your interpretation of the permutation test?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

The p-value for both the Sex factor and Training factor as well as for
the interaction 0, meaning that none of the permutations provides a
better result than the original. This can also be seen in the
permutation plots. The conclusion is that the Sex levels (Male and
Female) are different from eachother and also the training group levels
and that the interactions are not 0, and thus have to be included in the
model.
:::

Finally we can make a plot of the combination Training + Sex:Training to
visualize how the Females and Males deviate in their training effect.
For this we:

1.  Make ASCA model in which the Sex:Training interaction effect is
    combined with the main Training effect.
2.  Determine the explained variance of the components in combination
    effect.
3.  Make a score plot (using the timeplot function) of the combination
    effect. The PC1 scores of the Females and Males are plotted as a
    function of the training groups.
4.  Make a loading plot for PC1 of the combination effect.
5.  Also show a plot of only the interaction.


```{r }
#1. Make Model with main factor Sex and combination of Group and Sex:Group interaction
mod_comb1 <- asca(Metabolites ~ Sex + comb(Training+Sex:Training), data=Data_Values_long)
summary(mod_comb1)

#2.  determine explained variance of the combination effect
Var_Exp_Comb1 = 100*(mod_comb1[["singulars"]][["Training+Sex:Training"]]^2/sum(mod_comb1[["singulars"]][["Training+Sex:Training"]]^2))
Var_Exp_Comb1= round(Var_Exp_Comb1,0)

#3.
timeplot(mod_comb1, factor="Sex", time="Training", comb=2, comp=1,  x_time=TRUE, ylab=paste0("PC1 (", Var_Exp_Comb1[1], "%)"))
legend("topright",col = 1:2,lty=1:2,levels(Data_Values_long$Sex))

#4.
loadingplot(mod_comb1, factor = "Training+Sex:Training", comp=1,ylab=paste0("PC1 (", Var_Exp_Comb1[1], "%)"), main = ("Loadings Combination (Training+Sex:Training)"))
#5
Var_Exp_Int_Int= round(mod_asca_Int_perm[["singulars"]][["Sex:Training"]],0)
loadingplot(mod_asca_Int_perm, factor = "Sex:Training", comp=1,ylab=paste0("PC1 (", Var_Exp_Int_Int[1], "%)"),main = ("Loadings Interaction Sex:Training"))
```


::: {.callout-tip icon="false"}
## Question: What is your interpretation of the combination effect (Training + Sex:Training), and of the interaction Sex:Training
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer

-   The combination is significant (both factor Training and
    Sex:Training are significant).
-   Its scoreplot shows a Training time trend for each Sex.
-   Variation between the Training groups is somewhat larger for
    females than for males.
-   The metabolite levels change rapidly between 2 weeks and 4 weeks of
    training.
-   Note the differences between the loading plots for the combination
    and the interaction. The loading plot for the interaction only
    focuses on the difference between Females and Males, while the
    combination loadings combine the Main training variation with this
    difference between Females and Males. This results in different important features.
:::
