---
toc: true
toc-expand: 2
---

::: {.callout-note}
## Credits & copyright
BDA ASCA
author: "Johan Westerhuis"
:::

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Install and load the necessary packages

ASCA will be applied using the HDANOVA package, which also needs the mixlm package.

```{r}
# One-time installation

if (!requireNamespace("mixlm", quietly = TRUE)) {
	devtools::install_github("khliland/mixlm")
}
if (!requireNamespace("HDANOVA", quietly = TRUE)) {
	devtools::install_github("khliland/HDANOVA")
}

# in newer version of R you can also use pakage manager pak to install these packages
# pak::pak("khliland/mixlm")
# pak::pak("khliland/HDANOVA")

# load the libraries
library(mixlm)
library(HDANOVA)
```

# Exercise during lecture

```{r}
# create a simple dataset with 3 groups and 12 samples

# create response variable y and two different design matrices for the same grouping, one with reference coding and one with sum coding
y = c(5,6,7,6,3,2,4,3,1,2,3,1)
X0 = c(1,1,1,1,1,1,1,1,1,1,1,1) # not needed in calculation; Intercept is automatically calculated in lm function

# create design matrices for reference coding and sum coding
Xr1 = matrix(c(0,0,0,0,1,1,1,1,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1),ncol = 2) # 
Xs1 = matrix(c(-1,-1,-1,-1,1,1,1,1,0,0,0,0,-1,-1,-1,-1,0,0,0,0,1,1,1,1),ncol = 2)

# make linear models with reference coding and sum coding
bref = lm(y~Xr1)
bsum = lm(y~Xs1)

# look at the summary of both models
summary(bref)
summary(bsum)
```

::: {.callout-tip icon="false"}

## Question: What are the differences in interpretation?
What are the estimates for $\mu$, $\alpha_1$, $\alpha_2$, $\alpha_3$ in both cases?

:::

::: {.callout-caution collapse="true" icon="false"}
## Answer
* For Reference coding $\mu = 6$, $\alpha_1 = 0$, $\alpha_2 = -3$, $\alpha_3 = -4.25$. 
* Reference coding forces the intercept to be equal to the groupmean of group 1. The a1 value is 0 as group 1 is the reference group. 
* For Sum coding $\mu = 3.5833$, $\alpha_1 = 2.4166$, $\alpha_2 = -0.5833$, $\alpha_3 = -1.8333$. 
* Sumcoding makes the intercept to be equal to the average of all groups. The interpretation of the effects are therefore different between Reference and Sum coding. 
* Note however, that the full model statistics are exactly equal.
:::

# Brassicaceae study

In this exercise you will use the **HDANOVA** and **mixlm** packages to explore linear models and ASCA for analysis of high dimensional data from an designed study setup.

Plants in the Cabbage family (Brassicaceae) make very specific defense compounds, called glucosinolates, when under attack by herbivores. There are about 120 different versions of these glucosinolates, but here we measure only 11. To simulate a herbivore attack and elicit the plant to make glucosinolates, the plant hormone jasmonic acid (JA) was applied to either the roots (Root induced) or to the leaves (Shoot induced) of cabbage plants (B. oleracea). The glucosinolate levels for 11 different glucosinolates were measured at 1, 3, 7 and 14 days after treatment. This measurement was destructive so that for each time-point different plants were analyzed. For each treatment, 5 different plants were measured at each time point. Besides the two treatments there was also a control group that received no JA treatment. We would like to know how the different treatments affect the glucosinolate composition and whether the glucosinolate composition varies with time.

```{r}
# load data and design
# Data = read.csv("https://biodsc.nl/workshop-materials/bda-anova/Data.csv")
# Design = read.csv("https://biodsc.nl/workshop-materials/bda-anova/F.csv")

Data = read.csv("Data.csv")
Design = read.csv("F.csv")
```

In the Environment window (top right) click on Data and Design to see the content of these two sets. Data contains the measurements, and Design the experimental settings of the design.


::: {.callout-tip icon="false"}

## Question: Is the design balanced?
:::
::: {.callout-caution collapse="true" icon="false"}

## Answer
Yes, each combination of Treatment and Time has 5 replicates.
:::

::: {.callout-tip icon="false"}
## Question: What are important specifics you can observe by a quick look at the Data?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer
Some variables have many zeros, some variables have much higher values than others.
:::

## Data Preprocessing

First make a boxplot of the data to see which preprocessing steps should be taken.

```{r}
boxplot(Data)
```

::: {.callout-tip icon="false"}
## Question: Wat problematic issues do you observe for which variables?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
| Some features have few extreme large values (e.g. Neo, GBC), their distribution seems skewed to the right. Others have zero's (feature 2, 6, 9 etc).  
Sizes of the different features are also different (GBC and NEO are large).
:::

In such a case a square root transform is often a good choice to preprocess the data.

::: {.callout-tip icon="false"}
## Question: Explain why a square root transform is a useful transformation for these data.
:::


::: {.callout-caution collapse="true" icon="false"}
## Answer
| If there is more variation between the replicates for higher values of a feature, then a power transform can correct that. 
If there are zero values, a log transform is problematic. A square root transform does not have problems with zero values, and makes the distributions more symmetrix around the mean. A square root transform also makes the sizes of the variable more comparable.
:::

Make a data frame containing factors Time and Treatment and Data.

```{r}
# Make a data frame containing factors Time and Treatment and Data
dat = data.frame(sqrt(Data),Design);
dat$Time = as.factor(dat$Time)
dat$Treatment = as.factor(dat$Treatment)
```

## Linear models and ANOVA

To explore the effect of the treatments on the glucosinolates, let's start with an ANOVA model for a single response using the lm function. By default the lm function uses SUM coding for the treatment levels with the Control level as the one not calculated.

```{r}
# make 1-way ANOVA model for PRO using Treatment as a factor
mod_Treatment = lm(PRO~Treatment, data = dat)
print(anova(mod_Treatment))
```

There is clearly a difference between the treatments as the p-value is very low.
Get a little more info on the treatment levels, and look at the residual plots.

```{r}
# Get a little bit more information
summary(mod_Treatment)
plot(mod_Treatment, which = 1:2)
```

::: {.callout-tip icon="false"}
## Question: Explain why the effect of one of the treatment levels is not calculated. How can you obtain information about this level?
:::

::: {.callout-caution collapse="true" icon="false"}
## Answer
If all levels are represented in the design matrix, and the intercept is estimated as well, then there are too many free parameters leading to a situation with infinite number of solutions. By forcing the extra restriction that all treatment effect add up to 0 (which is the result of Sum coding) this situation is solved. The shoot level effect, which is not estimated, thus equals 0.05008.
:::

::: {.callout-tip icon="false"}
## Question: What is your interpretation of this model? What can you say about the root, shoot and control levels?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* The intercept is the average value of all measurements. 
* The root treatment provides PRO levels on average 0.24 lower than the average, while the shoot treatment provides PRO levels 0.19 higher than the average. 
* The control level is not indicated, but the sum of all treatment levels equals zero, thus the control level provides PRO levels around 0.05 higher than the average. 
* The root and shoot levels are clearly different from zero in the t-test. 
* The residual vs Fitted plot shows residuals are nicely spread around zero for all fitted values (which are 1.12 - 0.24 for control; 1.12+0.05 for control and 1.12+0.19 for shoot). 
* The QQ residuals plot to check normality shows three samples with slightly higher residuals than expected.
:::

### Two-way crossed ANOVA
Instead of a single factor we also want to know the effect of both factors and also their interaction.

```{r}
# two-way crossed ANOVA
mod <- lm(GBN ~ Time*Treatment, data = dat) # Note the * means all main effects and interactions
print(anova(mod))
```
For GBN, we see that both factors (Time and Treatment) are significant as is their interaction.
Also here check the significance of levels of the factors.

```{r}
summary(mod)
plot(mod, which = 1:2)
```
::: {.callout-tip icon="false"}
## Question: Explain your interpretation of this model.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* For GBN there is a clear treatment effect and also some timepoints are different. 
* There is a single interaction clearly different from zero. 
* The GBN values at time 3 for the control treatment are 0.17 higher than could be expected from their main factor results control treatment effect and day 3 effect.
:::

Different codings can be used for the factors. Here you can compare sum coding and treatment (reference) coding.

```{r}
# Different codings
mod_sum <- lm(GBN ~ Time*Treatment, data = dat, contrasts = "contr.sum")
print(anova(mod_sum))
summary(mod_sum)

mod_treat <- lm(GBN ~ Time*Treatment, data = dat, contrasts = "contr.treatment")
print(anova(mod_treat))
summary(mod_treat)
```

::: {.callout-tip icon="false"}
## Question: Discuss the differences between the two types of coding. When is one preferred over the other? 
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
The difference is mainly what is considered the intercept (base level) of the model. If a control level is present, one would like to see how the treated level is different from the control instead of how the both differ from their average.
:::
::: {.callout-tip icon="false"}
## Question: What is the interpretation of the intercept in the models? What is the meaning of the estimated coefficients in each model?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
In the SUM coded model, the intercept represents the overall mean, but in the treatment coding the intercept represents the mean of the Control treatment samples at time 1. Check this! The effects of the other levels mean the difference between those levels and the Control at time 1 level.
:::

## ASCA

ASCA generalized ANOVA or the linear model to the multivariate setting. The models are applied to all response variables, and the variation due to the different factors (and their interactions) is summarized into effect matrices. Then a PCA model can be applied to each effect matrix separately of in combination to find variation that is systematic over all the variables. This variation is then summarized with PCA scores and loadings.

We will use the **HDANOVA** package to apply asca. First make data from wide to long format for the asca function.

```{r}
# Reshape data from wide to long format
Metabolites = data.frame(Time = dat$Time,Treatment = dat$Treatment, Metab = I(as.matrix(dat[,1:11])))
Metabolites$Time = as.factor(Metabolites$Time)
Metabolites$Treatment = as.factor(Metabolites$Treatment)
```

Then make an ASCA model of the metabolite data using both Time and Treatment as factors together with their interaction.

```{r}
# ASCA model of all variables simultaneously
mod_asca <- asca(Metab ~ Time*Treatment, data=Metabolites)
summary(mod_asca)
```

Open de mod_asca list and look at all the parameters that have been calculated. Click the scores section. The scores are calculated For each of the effect matrices. This is also the case for the loadings and Projected scores. The latter are the score values calculated with back projection, using the residuals and the effect matrices together, as discussed in the lecture.

```{r}
# Simple functions to use
scoreplot(mod_asca, factor="Treatment", legend = ("topright"), main="Treatment scores")
loadingplot(mod_asca, factor="Treatment", scatter=TRUE, labels="names", main="Treatment loadings")

# For more control you can obtain information from the mod_asca list, e.g. when you only want to plot the top 8 loadings
# SCoreplot
T_Treatment = mod_asca[["scores"]][["Treatment"]];
Var_Exp_Treatment = 100 * mod_asca[["singulars"]][["Treatment"]]^2 / sum(mod_asca[["singulars"]][["Treatment"]]^2)
Var_Exp_Treatment = round(Var_Exp_Treatment,0);
colours = as.numeric(as.factor(Metabolites$Treatment));
plot(T_Treatment[,1],T_Treatment[,2], pch = 16, col=as.factor(Metabolites$Treatment),xlab=paste0("PC1 (", Var_Exp_Treatment[1], "%)") , ylab=paste0("PC2 (", Var_Exp_Treatment[2], "%)") , main="PCA of Treatment effect" )
legend("topright", legend=unique(as.factor(Metabolites$Treatment)), col=unique(colours), cex=0.8, pch=16, title="Treatment")

# Loading plot
P_Treatment = mod_asca[["loadings"]][["Treatment"]];
Importance = P_Treatment[,1]^2 + P_Treatment[,2]^2;
NIMP = 8; # number of important variables
sorted_Importance = sort(Importance, decreasing = TRUE, index.return = TRUE)
Importance_Index = sorted_Importance$ix[1:NIMP]
plot(P_Treatment[Importance_Index,1],P_Treatment[Importance_Index,2],type = "n",xlab=paste0("Comp 1 (", Var_Exp_Treatment[1], "%)") , ylab=paste0("Comp 2 (", Var_Exp_Treatment[2], "%)") )
text(P_Treatment[Importance_Index,1],P_Treatment[Importance_Index,2],colnames(Data[,Importance_Index]))
```
::: {.callout-tip icon="false"}
## Question: What is your interpretation of the ASCA model of the Treatment factor?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* There is a clear difference between the shoot and control treatments, which are described by the 1^st^ PC of the treatment factor. GBC and NEO have the largest contribution to this effect. 
* The 2^nd^ PC which is much smaller wrt explained variation shows how the root treatment is different from control and shoot, and here GBN and PRO have the largest contribution. Check this by looking in the data whether this observation is correct!
:::

It is important to check whether the grouping of the factors is significant. This can be done using permutation testing.

```{r}
mod.perm <- asca(Metab ~ Time*Treatment, data=Metabolites, permute = TRUE)
summary(mod.perm)
par(mfrow=c(1,3))
permutationplot(mod.perm, factor = "Time")
permutationplot(mod.perm, factor = "Treatment")
permutationplot(mod.perm, factor = "Time:Treatment")
```

::: {.callout-tip icon="false"}
## Question: What conclusion can you draw from the permutation test results?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
Both factors as well as the interaction have a significant effect on the glucosinolate levels. For the main effects none of the permutations gave better results than the true grouping, while for the interaction only few permutations were larger than the true grouping.
:::

#### Combining Main and interaction effects
Sometimes it makes sense to combine effects, e.g., a main factor and an interaction. This is especially the case when the effect of Treatments are studied over Time. Here we combined the effects of Treatment and the Time:Treatment interaction. Note the effect matrices of Treatment and Time:Treatment are estimated independently, but the PCA is applied on the sum of these two effect matrices.

```{r}
mod.comb <- asca(Metab ~ Time + comb(Treatment + Time:Treatment), data=Metabolites)
summary(mod.comb)
```

If combined effects have a Time factor, the scores can be plotted against the time using the timeplot function. For this plot the legend command has to be called separately.

```{r}
par(mfrow=c(1,2), mar = c(4,4,1,1))
timeplot(mod.comb, factor="Treatment", time="Time", comb=2, comp=1,  x_time=TRUE)
legend("topright",col = 1:3,lty=1:3,levels(Metabolites$Treatment))
loadingplot(mod.comb, factor = "Treatment+Time:Treatment", comp=1)
```

::: {.callout-tip icon="false"}
## Question: Give your interpretation of this combined effect.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
In the scoreplot we see the combined effect of treatment and time. This shows that the effect of treatments is not constant over time. The biological explanation is that both root and shoot treatments have an effect, but after day 3, the root effect dissappears and the glucosinolate levels go back to the control values. The shoot effect however is consistent and even increases over time. Variables 8 and 11 are largely effected by these treatment differences.
:::

Finally, it can be useful to also explore the residual matrix.

```{r}
# Make score and loading plots and interpret.
par(mfrow=c(1,2)) 
loadingplot(mod.comb, factor="Residuals", scatter=TRUE, labels="names", main="Residual loadings")
scoreplot(mod.comb, factor="Residuals", scatter=TRUE, main="Residual scores")
```
::: {.callout-tip icon="false"}
## Question: Why is it useful to also explore the residuals and interpret the PCA model of the residuals.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
We can see that some plants (the cluster right bottom deviate from the main group. These will have higher values of NEO and lower values of GBC than the other plants.)
:::


# Additional exercises
In this 2^nd^ exercise you have to do some coding yourself. 

### The MoTrPAC data

The Molecular Transducers of Physical Activity Consortium (MoTrPAC)
profiled the dynamic change of multiple omics layers in several organs
of rats undergoing physical training. The results are presented in the
Nature article "Temporal dynamics of the multi-omic response to
endurance exercise training". [https://www.nature.com/articles/s41586-023-06877-w](https://www.nature.com/articles/s41586-023-06877-w)
The data was obtained from their website MoTrPAC-data.org.

### Experimental design

Six-month-old male and female Fischer 344 rats were subjected to
progressive treadmill endurance exercise training for 1, 2, 4 or 8
weeks. Sex-matched sedentary, untrained rats were used as controls. In
this application we will focus on metabolomics data from plasma. 
Unfortunately these measurements do not all come from the same rats.
Note that Females have Sex = 1 and Males have Sex = 2.

#### Load and view the data
Load the metabolomics_Values (Data) and the metabolomics_Exp (Design) files.
For visualization reasons we will only select the lipids subset.

As preprocessing we decided to not use a transformation of the data, 
but only use meancentering and scaling by the standard deviation of the lipid.

```{r Assignment 6A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# load data and design
# df_metabolomics_Values = read.csv("https://biodsc.nl/workshop-materials/bda-anova/Plasma_Metabolomics_Values_a.csv", row.names=1)
# df_metabolomics_Exp = read.csv("https://biodsc.nl/workshop-materials/bda-anova/Plasma_Metabolomics_Exp_a.csv", row.names=1)
df_metabolomics_Values = read.csv("Plasma_Metabolomics_Values_a.csv", row.names=1)
df_metabolomics_Exp = read.csv("Plasma_Metabolomics_Exp_a.csv", row.names=1)


df_metabolomics_Lipids = df_metabolomics_Values[,80:613]

# make boxplots of the data before preprocessing
par(mfrow=c(1,3)) 
boxplot(df_metabolomics_Lipids)

# Preprocess the data by centering and scaling 
CData = scale(df_metabolomics_Lipids, center = TRUE, scale = FALSE)
SData = scale(df_metabolomics_Lipids, center = TRUE, scale = TRUE)

# Make boxplots of the data after centering and scaling
boxplot(CData)
boxplot(SData)
```
::: {.callout-tip icon="false"}
## Question: Why did we preprocess the lipids data this way?
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* From the 1^st^ boxplot we see that the Lipids data is not centered.
* We can also see the distributions are not skewed but rather symmetrical, thus transformation is not necessary. 
* Scaling seems necessary as the ranges of the centered data are rather different for some metabolites. 
* The variation within samples is rather different between metabolites. Therefore, transformation does not seem to be important, but centering and scaling the data is important.
:::

#### Data handling
Just as above the data has to be made in the long format to be able to use the **HDANOVA** package. 
Make sure the factors Sex and Group are defined as factors. 

```{r Assignment 6B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# make the data frame in long format
Data_Values_long = data.frame(Sex = as.factor(df_metabolomics_Exp$Rat.Sex), Group = as.factor(df_metabolomics_Exp$Group), Metabolites = I(as.matrix(SData)))
```

#### ASCA model of main factors: 
Create an ASCA model using the only the main factors
Sex and Group. Do not consider interaction terms yet.

```{r Assignment 6C, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# ASCA model of main factors
mod_asca <- asca(Metabolites ~ Sex + Group, data=Data_Values_long)
# Look at the summary of the model
summary(mod_asca)
```

::: {.callout-tip icon="false"}
## Question:Explain the meaning of Sum.Sq and Expl.var. What is your interpretation of the ASCA model? 
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* Sum.Sq is the sum of squares of the effect matrix, calculated by squaring each element in the effect matrix and then adding all these values
* Expl.var is the percentage explained variation from the total variation in the Metabolites data. It is calculated by dividing the Sum.Sq of a factor (e.g. Sex) by the Sum.Sq of the original data (Metabolites)
* There is a strong effect for factor Sex and a smaller effect for factor Group
:::

#### Visualization
1) Make score plot
2) Show the loading plots of PC1 vs PC2 of the Group factor. Only show the 20
most important variables where
$Importance_j =\sqrt(p_{1j}^2 + p_{2j}^2)$. [1.5] 

::: {.callout-tip icon="false"}
## Question: Show score plot of PC1 vs PC2 of the Group factor. 
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
```{r}
# Make score plot.
scoreplot(mod_asca, factor="Group", legend = ("topright"), main="Group scores")
```
:::

Below the loading plots of PC1 vs PC2 of the Group factor is given.
For visualization reasons we only show the 20 most important variables.

```{r} 
# Loading plot of only 20 most important metabolites
# determine explained variance per principal component 
Var_Exp_Group = 100*(mod_asca[["singulars"]][["Group"]]^2/sum(mod_asca[["singulars"]][["Group"]]^2))
Var_Exp_Group = round(Var_Exp_Group,0)
# extract loadings of the Group factor
P_group = mod_asca[["loadings"]][["Group"]];
# combine the loadings of PC1 and PC2 into an importance value
Importance = sqrt(P_group[,1]^2 + P_group[,2]^2);
NIMP = 20; # number of important variables
# sort the importance values and get the index of the top 20 most important variables
sorted_Importance = sort(Importance, decreasing = TRUE, index.return = TRUE)
Importance_Index = sorted_Importance$ix[1:NIMP]
# plot the top20 most important variables in the loading plot
plot(P_group[Importance_Index,1],P_group[Importance_Index,2],xlab=paste0("Comp 1 (", Var_Exp_Group[1], "%)"), ylab=paste0("Comp 2 (", Var_Exp_Group[2], "%)"))
text(P_group[Importance_Index,1],P_group[Importance_Index,2],colnames(df_metabolomics_Lipids[,Importance_Index]))
```

::: {.callout-tip icon="false"}
## Question: Give your interpretation of the score and loading plots. [1]
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
The scoreplot shows the group average of the 5 groups by the solid marker, while the individual rats are indicated with the open markers. The 4 and 8 weeks training samples are mainly in the top left part of the plot, well separated from the other samples. Component 1 mainly makes a distinction between Training week1 and week2.
:::

**The extreme loading values are found upper left and lower right. For example, the object PC.42.8 has high values for the 1 and 2 weeks training and are low for week4 and week8 training**

#### ASCA model with interaction

::: {.callout-tip icon="false"}
## Question: Make a new model in which the two-factor interaction Sex:Group is also included in the model.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
```{r}
mod.perm <- asca(Metabolites ~ Sex + Group + Sex:Group, data=Data_Values_long)
summary(mod.perm)
```
:::
::: {.callout-tip icon="false"}
## Question: What is your conclusion of this ASCA model with interaction?
:::
::: {.callout-caution collapse='true' icon='false'}
## Answer
| The interaction explains an additional 7.7% of the variation in the data. 
This means that the Time effect is somewhat different for males compared to females.
:::

#### Statistical Inference 
To conclude whether the main factors and interaction are significant, apply a permutation test and plot the
permutation result using the permutation plot function for both main factors and the interaction.

```{r}
# Permutation test for the ASCA model with interaction
mod.perm <- asca(Metabolites ~ Sex + Group + Sex:Group, data=Data_Values_long, permute = TRUE)
summary(mod.perm)
# make permutation plots for the main factors and the interaction
par(mfrow=c(1,3))
permutationplot(mod.perm, factor = "Sex")
permutationplot(mod.perm, factor = "Group")
permutationplot(mod.perm, factor = "Sex:Group")
```
::: {.callout-tip icon="false"}
## Question: Explain why in ASCA a permutation test is used and not an F-test.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
The F-statistic assumes normally distributed data and also requires the degrees of freedom for each effect. We have not tested the normality of the data, and degrees of freedom is unclear in the multivariate situation.
:::

::: {.callout-tip icon="false"}
## Question: What is your interpretation of the permutation test? 
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
The p-value for both the Sex factor and also for the combination (Group+Sex:Group) is 0, meaning that none of the permutations provides a better result than the original. This can also be seen in the permutation plots. The conclusion is that the Sex levels (Male and Female) are different from eachother and also the combination levels (Group+Sex:Group) are different from each other.
:::



When exploring the model with interaction, it is common to combine the interaction effect with one of the main factor effects to improve
interpretation. 

::: {.callout-tip icon="false"}
#### Questions:
1. Make ASCA model in which the Sex:Group interaction effect is combined with the main Group effect. 
2. Determine the explained variance of the components in combination effect.
3. Make a score plot (using the timeplot function) of the combination effect. The PC1 scores of the Females and Males are plotted as a function of the training Groups. 
4. Make a loading plot for PC1.
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer

```{r Assignment 6G ASCA Combination model, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
#1. Make Model
mod.comb1 <- asca(Metabolites ~ Sex + comb(Group + Sex:Group), data=Data_Values_long)
summary(mod.comb1)

#2.  determine explained variance of the combination effect
Var_Exp_Comb = 100*(mod_asca[["singulars"]][["Group + Sex:Group"]]^2/sum(mod_asca[["singulars"]][["Group + Sex:Group"]]^2))
Var_Exp_Comb= round(Var_Exp_Comb,0)

#3.
timeplot(mod.comb1, factor="Sex", time="Group", comb=2, comp=1,  x_time=TRUE, ylab=paste0("PC1 (", Var_Exp_Comb[1], "%)"))
legend("topright",col = 1:2,lty=1:2,levels(Data_Values_long$Sex))

#4.
loadingplot(mod.comb1, factor = "Group+Sex:Group", comp=1,ylab=paste0("PC1 (", Var_Exp_Comb[1], "%)"))
```
:::

::: {.callout-tip icon="false"}
## Question: What is your interpretation of the combination effect (Group + Sex:Group)
:::
::: {.callout-caution collapse="true" icon="false"}
## Answer
* The combination is significant (both factor Sex and Sex:Group are significant). 
* Its scoreplot shows a timetrend for  line for each group how the score value is different for males and females. 
* Variation between the groups is somewhat larger for females than for males. 
* The metabolite levels change rapidly between 2 weeks and 4 weeks of training. 
* Note also that rats also get older which might change their metabolome and that Control 8 weeks is different from training 0 weeks.
:::